\documentclass[main.tex]{subfiles}

\begin{document}

\chapter{An Analog to Snyder's Parameter}

Here begin the main results of this thesis. The following chapter is devoted to a group theoretic analog to Snyder's parameter. We replace his use of irreducible character degrees with the square roots of conjugacy class sizes. Both of these satisfy the identity:
$$|G| = \sum_{i=1}^n d_i^2$$
where $d_i$ is either character degrees or square roots of conjugacy class sizes and $i$ indexes the total set of characters or conjugacy classes (the number of objects indexed is the same in both cases (cite)). After initial definitions and examples, we will analyze the properties of and range of values $e$ can take. Then, we will bound the order of a group by our parameter and characterize groups attaining this bound.

Without further ado:

\hss

\begin{definition}
Let $G$ be a finite group. For $x \in G$ let $c(x) = |C_G(x)|$ and $k(x) = |G : C_G(x)|$. We define the parameter $e$ as follows:
$$e = min\{(c(x) - 1) \cdot \sqrt{k(x)} : x \in G\}\text{.}$$
\end{definition}

\hss

To  see that this really corresponds with Snyder's $e$, let $x \in G$ be such that $e = (c(x) - 1) \cdot \sqrt{k(x)}$ and consider the following:
\begin{align*}
e &= (c(x) - 1) \cdot \sqrt{k(x)} \\
 &= c(x)\frac{k(x)}{\sqrt{k(x)}} - \sqrt{k(x)} \\
 &= \frac{|G|}{\sqrt{k(x)}} - \sqrt{k(x)}\text{.}
\end{align*}
If we let $d = \sqrt{k(x)}$, we arrive at the equation $|G| = d (d + e)$. Since $k(x)$ is precisely the size of the conjugacy class containing $x$, for a fixed $G$, when the conjugacy class of $x$ is very large, $e$ is very small. This behavior is similar to Snyder's parameter with character degrees. 

For a more detailed analysis of this relation, consider the function $f(c) = (c - 1) \sqrt{n/c}$ where $n$ is the order of a group and $c$ is the order of a centralizer. Then $e$ is the minimum of $f$ over the domain of possible centralizer orders. Since $f'(c) = \sqrt{n/c} \cdot (c+1) / 2c$ and $c \ge 1$, the function $f$ must be monotonically increasing. Thus, small values of $e$ are correlated with small centralizers or, equivalently, large conjugacy class sizes. In particular, the smallest possible values of $e$ are attained only by groups with a conjugacy class that contains exactly half of the group's elements, the largest non-trivial conjugacy class possible relative to a group's order (see Theorem \ref{boundaryclassification}).

For abelian groups, calculating $e$ is nearly trivial. Since all conjugacy classes are singletons in such groups, we have $e = n - 1$. Another easy example is the the symmetric groups. The largest conjugacy class in $S_n$ is that of the $(n-1)$-cycles (cite) and it has size $n!/(n-1)$. The value of $e$ for $S_n$ is then:
$$e = (n - 2) \cdot \sqrt{\frac{n!}{n-1}}\text{.}$$

\hss

We now analyze the smallest possible values our parameter $e$ can take. Note that the groups here are very small and that $e$ necessarily increases for larger groups (a closer look at the function $f$ mentioned above reveals that $f$ increases as $n$ does). In fact, given the order of a group, we can specify the range of values $e$ can take. We need only look at the largest possible conjugacy class and the smallest possible conjugacy class. If $G$ is a non-trivial group, it's largest conjugacy class can contain up to half of the group's elements. This case corresponds to a centralizer of order 2. For the other extreme, a conjugacy class can contain only a single element. This happens in abelian groups and the entire group is the centralizer for every element. Then, for a group $G$ of order $n$:
\begin{equation}\label{erange}
\sqrt{\frac{n}{2}} \le e \le n - 1\text{.}
\end{equation}
This fact makes it easy to classify groups with small values of $e$.

\begin{theorem}\label{easycharacterization}
Let $G$ be a finite group.
\begin{enumerate}
	\item $e = 0$ if and only if $G$ is trivial.
	\item $e = 1$ if and only if $G$ is $C_2$.
	\item $e = 2$ if and only if $G$ is $C_3$.
\end{enumerate}
\end{theorem}

\begin{proof}
For the following, let $x \in G$ be an element such that $e = (c(x) - 1) \cdot \sqrt{k(x)}$.

Suppose $e = 0$. Then $c(x) = 1$. Only the trivial group has a trivial centralizer. This is sufficient for (1).

Let $e = 1$. Then $c(x) = 2$ and $k(x) = 1$. It must be the case that $|G| = 2$ and so $G$ is $C_2$. Reversing this argument proves (2).

Now, let $e = 2$. There are two cases. If $c(x) = 3$ and $k(x) = 1$, then $G$ must be $C_3$. Since $C_3$ must have $e = 2$, we need only consider the remaining case. Assume $c(x) = 2$ and $k(x) = 4$. This implies that $|G| = 8$. Since $c(x) < 8$, $G$ cannot be abelian. The only two groups of order 8 that are not abelian are $D_8$ and $Q_8$ \cite{smallgroups}, the dihedral group of order 8 and quaternion group respectively. Neither of these groups have a self-centralizing involution and thus $G$ can be neither. This completes (3).
\end{proof}

Due to equation (\ref{erange}), if a group is non-trivial, $e \ge 1$. This gives, as a nice corollary, that $e$ cannot take on any value between 0 and 1. The parameter $e$ can, however take on values between $1$ and $2$. Our work with the symmetric group earlier in the chapter showed that $S_3$ has $e = \sqrt{3}$ in this range. Interestingly, this is the only exception. The largest possible group that can still have an $e < 2$ is of order 7 since $\sqrt{7/2} < 2$ but $\sqrt{8/2} \not< 2$. There are no non-abelian groups of order 3, 4, 5, or 7 \cite{smallgroups} so any group that would have $1 < e < 2$ must be of order 6. There are only two groups of order 6, the cyclic group of order 6 and $S_3$. Since the former is abelian, $S_3$ is the only group with an $e$ between 1 and 2.

\hss

Next, we extend the work above and classify groups with prime values of $e$. This result is the beginning of a complete collection of understood $e$ values. Even though $e$ is not always an integer, it is based on integral values and, hence, does not range over all real numbers (as we saw above, $e$ takes exactly one value between 1 and 2). Furthermore, this theorem illustrates that $e$ has a strong relationship with the structure of a group. Before we get to this classification, we present a required lemma.

\begin{lemma}\label{psquared}
Let $P$ be a group of order $p^2$ for some prime $p$. Then $P$ is abelian.
\end{lemma}

\begin{proof}
Let $P$ act on $Z(P)$, the group's center, by conjugation. We know that every element of $Z(P)$ has an orbit of size one under this action and every other orbit must have $p$-power size. Then $P - Z(G)$ is a union of orbits with size divisible by $p$. Thus, $|Z(P)| \equiv 0 \pmod{p}$ and, since $1 \in Z(P)$, $|Z(P)| \ge p$. 

If $Z(P) = P$ we are done, so suppose $|Z(G)| = p$. Then there is an element $x \in P - Z(P)$ and $\generate{x} \cap Z(P) = 1$. Therefore, the group $\generate{x}$ is  a complement of $Z(G)$ and $P = Z(G) \rtimes_\varphi \generate{x}$ for some $\varphi \in Aut(Z(P))$. But, for $z \in Z(P)$ we must have $z^x = z$. Then $\varphi$ is the identity and $P = Z(P) \times \generate{x}$. Since both $Z(P)$ and $\generate{x}$ are abelian, we must conclude that $Z(P) = P$.
\end{proof}

Now, we get to a more sweeping characterization:

\begin{theorem}
Let $G$ be a finite group and $p$ be an odd prime. The group $G$ is abelian of order $p+1$ or generalized dihedral of order $2p^2$ if and only if $e = p$.
\end{theorem}

\begin{proof}
Suppose that $G$ is abelian of order $p+1$. Then $c(x) = p+1$ for all $x \in G$. This gives that $k(x) = 1$ for all $x \in G$ and we may conclude that $e = (c(x) - 1) \cdot \sqrt{k(x)} = p$.

For the other case, let $G$ be generalized dihedral of order $2p^2$. Take $t \in G$ to be an involution. Since $G$ is generalized dihedral, the conjugacy class size of $t$ is $p^2$ by Proposition \ref{dihedralinvolutionclass} since $p$ is odd. This conjugacy class is as large as it can be and so corresponds to the minimal value of $e$. Then $e = (c(t) - 1) \cdot \sqrt{k(x)} = \sqrt{k(x)}$. But, $k(x)$ is the size of $t$'s conjugacy class, $p^2$, and so $e = p$.

All that remains is to be shown is that no other classes of groups have such an $e$. Let $e = p$. Since $e$ is an integer, we know that $k(x)$ must be a perfect square, less we arrive at a contradiction (this holds due to the elementary result that the product of a rational and irrational number must be irrational). Then either $c(x) = p+1$ and $k(x) = 1$ or $c(x) = 2$ and $k(x) = p^2$. In the former case, we know that the group is of order $p+1$. Since $e$ is minimal, we know that $c(x)$ is the size of the smallest centralizer. Then every element centralizes every other element and the group must be abelian. In the latter case, we know that the group must be of order $2p^2$ and have a self-centralizing involution. But, this implies that there are $p^2$ involutions in $G$. Furthermore, $G$ must have a Sylow $p$-subgroup $P$ and it must be disjoint from these involutions as $p$ is an odd prime. Since $P$ has prime-power order $p^2$, it must be abelian by Lemma \ref{psquared}. Then $G$ is generalized dihedral.
\end{proof}

This result is much stronger than any similar one for the character theoretic $e$. Snyder specifically mentioned in \cite{snyderarticle} that arguments classifying groups with particular values of $e$ did not generalize. That is not the case here.

\hss

It is possible to bound the order of a group by $e$. For Snyder's orginal parameter, much work has been done to make a better bound (\cite{isaacsarticle}, \cite{durfeejensenarticle}, \cite{lewisarticle}). In our case, the answer is much easier and the bound much better:

\begin{theorem}\label{ebound}
Let $G$ be a non-trivial finite group. Then $|G| \le 2e^2$.
\end{theorem}

\begin{proof}
We begin with the fact that $k(x) = |G|/c(x)$. This gives the following equation for some $x \in G$:
\begin{align*}
e &= (c(x) - 1) \cdot \sqrt{k(x)} \\
&= \frac{(c(x) - 1)}{\sqrt{c(x)}} \cdot \sqrt{|G|}
\end{align*}
If we square this equation we get:
$$e^2 = \frac{(c(x) - 1)^2}{c(x)} \cdot |G|$$
from which we may deduce that:
$$|G| = e^2 \cdot \frac{c(x)}{(c(x) -1)^2}\text{.}$$
Since $c(x)/(c(x) - 1)^2$ decreases as $c(x)$ increases, we need only look at the minimal value of $c(x)$. Our group $G$ is non-trivial and thus all of its centralizers have at least 2 elements. Then $c(x) \ge 2$. This yields the following inequality:
$$2e^2 \ge e^2 \cdot \frac{c(x)}{(c(x) - 1)^2} = |G|$$
which completes the proof.
\end{proof}

A natural follow-up to this theorem is the question: Which groups attain the bound? Before we answer this, let us a review the following basic fact:

\begin{lemma}[\cite{isaacsfinitegrouptheory}]\label{ordernormal}
Let $|G| = 2n$ for an odd positive integer $n$. Then $G$ has a normal subgroup of index 2. 
\end{lemma}

\begin{proof}
Consider $\pi: G \to Sym(G)$, the permutation representation of the right multiplication action of $G$ on itself. By Cauchy's Theorem, we can find an element $t \in G$ of order two. Under the aforementioned action, $t$ has no fixed points. So $\pi(t)$, the permutation of $Sym(G)$ such that $g \cdot \pi(t) = gt$ for $g \in G$, consists of $|G|/2 = n$ 2-cycles. Since $n$ is odd, $t$ is an odd permutation.

Let $H$ be the elements of $G$ that induce even permutations on $G$. Take $x \in G - H$. Since $x$ induces an odd permutation, $\pi(xt)$ is an even permutation. Then $xt \in H$ and so $x \in Ht$. Thus $Ht = G- H$. We conclude that $H$ has index 2 and must be normal.
\end{proof}

We now classify groups attaining the bound mentioned above.

\begin{theorem}\label{boundaryclassification}
Let $G$ be a finite group. The group $G$ is generalized dihedral of order not divisible by 4 if and only if $|G| = 2e^2$. 
\end{theorem}

\begin{proof}
Suppose $G$ is generalized dihedral with order not divisible by 4. Such a group has $|G|/2$ involutions that are all conjugate. Then there exists an $x \in G$ such that $c(x) = 2$. Since this is the largest possible conjguacy class, it corresponds to the smallest possible $e$. Then:
$$e^2 = \frac{(c(x) - 1)^2}{c(x)} \cdot |G| = \frac{1}{2} \cdot |G|$$
and so $|G| = 2e^2$.

Now, suppose a finite group $G$ has order $2e^2$. We know from the proof of Theorem \ref{ebound}, that this implies $c(x) = 2$ for some $x \in G$. 

Then $x$ has conjugacy class size of $e^2 = n / 2$. Also, $e^2$ cannot be divisible by 2 lest $x \in G$ be contained an abelian subgroup of order $p^2$ by Lemma \ref{psquared}.  Now, we are in the situation of Lemma \ref{ordernormal} and may conclude that $G$ has a normal subgroup $B$ of index 2. Since $x$ is conjugate to $e^2$ involutions (including itself) and none of these can be contained in $B$ (its order is not divisible by 2), $G - B$ must be exactly $x^G$. Thus, $G$ is generalized dihedral of order not divisible by 4.
\end{proof}

\end{document}